---
output: github_document
---

<!-- README.md is generated from README.Rmd. -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

<div class="gs-hero">
<p class="gs-kicker">R package for clinical, epidemiological, and public-health analysis</p>
<h1>gtstats</h1>
<p>Understand your data, build a publication-ready Table 1, and answer focused statistical questions without hand-formatting every output.</p>
<p><a class="gs-button" href="https://gtstats.thinkdenominator.com/articles/getting-started.html">Start the workflow</a> <a class="gs-button gs-button-secondary" href="https://gtstats.thinkdenominator.com/articles/birthweight-case-study.html">View the case study</a></p>
</div>

<!-- badges: start -->

[![R-CMD-check](https://github.com/ThinkDenominator/gtstats/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ThinkDenominator/gtstats/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/ThinkDenominator/gtstats/actions/workflows/pkgdown.yaml/badge.svg)](https://gtstats.thinkdenominator.com/)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/license/mit)
[![R: >= 4.1](https://img.shields.io/badge/R-%3E%3D4.1-276DC3?logo=r)](https://www.r-project.org/)

<!-- badges: end -->

## Publication-ready descriptive and inferential statistics

`gtstats` helps students, researchers, and public-health analysts move from a
new dataset to clear descriptive results, focused comparisons, and a
publication-ready Table 1. It keeps the R syntax approachable while retaining
the underlying results, denominators, assumptions, and automatic decisions for
review.

| Stage | Start with | What you get |
|---|---|---|
| Understand | `describe_data()` | Variable type, completeness, levels/range, and concise data overview |
| Assess | `assess_distribution()` | Distribution diagnostics and descriptive guidance for continuous variables |
| Assess | `assess_variance()` | Group SDs, variances, spread ratios, and median-centred Levene support |
| Describe | `summary_table()` | Publication-ready participant-characteristics table |
| Present calculated results | `as_stats_table()` | Value-preserving publication table for an already summarised data frame |
| Compare | `compare_groups()` | One focused group comparison with a clearly identified test and automatic-selection rule |
| Audit | `assumptions_stats()`, `diagnostics_stats()`, `denominators_stats()` | Transparent decisions and analysis population |
| Export | `customise_table()`, `to_gt()`, `save_output()` | A modifiable, report-ready table |
| Guided | `gtstats_app()` | A point-and-click companion that generates the matching R code |

<div class="gs-strip">
<div class="gs-strip-item"><p class="gs-strip-label">Understand</p><p class="gs-strip-fns">describe_data()<br>assess_distribution()</p></div>
<div class="gs-strip-item"><p class="gs-strip-label">Describe</p><p class="gs-strip-fns">summary_table()<br>add_*()</p></div>
<div class="gs-strip-item"><p class="gs-strip-label">Compare</p><p class="gs-strip-fns">compare_groups()<br>correlation()</p></div>
<div class="gs-strip-item"><p class="gs-strip-label">Report</p><p class="gs-strip-fns">customise_table()<br>save_output()</p></div>
</div>

## Why it exists

Statistics packages often make a user choose between an oversimplified default
and a highly technical workflow. `gtstats` is designed to bridge that gap:

- publication-ready tables are the default, not an extra formatting task;
- automatic test selection is visible and can always be overridden;
- Welch methods are the conservative default for suitable independent continuous
  comparisons; `var_equal = TRUE` is available when equal variances are
  prespecified and justified, not inferred from a variance test;
- distribution, variance, assumptions, and denominators are inspectable;
- the table builder offers simple defaults with optional layers of control.

The automatic algorithm is documented in full: marked skewness—not a
Shapiro-Wilk p-value alone—triggers rank-based continuous tests; sparse
categorical tables use Fisher's exact test when an expected count is below 1
or more than 20% are below 5; and paired/repeated designs use their dedicated
methods. See the [inferential tests and assumptions
guide](https://gtstats.thinkdenominator.com/articles/inferential-tests.html).

`gtstats` uses established R statistical engines such as `stats`, `dplyr`,
`ggplot2`, `gt`, and `flextable`; it provides a coherent teaching and reporting
workflow around them.

## Install

```r
# Development version until gtstats is available on CRAN
remotes::install_github("ThinkDenominator/gtstats")
library(gtstats)
```

## Built-in teaching data

gtstats includes five labelled datasets, so examples work without installing
other packages:

```r
data(
  "birthwt", "trial_data", "paired_data", "outbreak_data",
  "surveillance_data", package = "gtstats"
)
```

- `birthwt`: real low-birth-weight clinical data for descriptive tables,
  independent comparisons, and epidemiology.
- `trial_data`: a three-arm synthetic trial for Welch ANOVA,
  Kruskal-Wallis, chi-square, Fisher exact, correlations, and rates.
- `paired_data`: long-format paired follow-up data for paired t-tests,
  Wilcoxon signed-rank, and McNemar tests.
- `outbreak_data`: the real CDC Oswego foodborne-outbreak line list for attack
  rates and 2x2 epidemiology.
- `surveillance_data`: an archived CDC weekly hospital-admission extract for
  aggregate counts, population denominators, and rates per 100,000.

## Prefer a guided interface?

The package remains code-first, but a Shiny companion is available when you
prefer to learn by clicking through the workflow. It includes all five
teaching datasets, data from the current R environment, CSV/Excel upload, a
working data dictionary, distribution and variance checks, a layer-by-layer
summary table, outbreak and surveillance tables, focused group comparisons,
correlations and crosstabs. Tables can be downloaded as Word, HTML, PDF, or
RTF, and every analysis has copyable and downloadable R code. Excel import
uses the optional `rio` package.

```r
install.packages("shiny") # once, if needed
install.packages("rio")   # once, only for Excel files
gtstats_app() # opens in the RStudio Viewer when available
```

The interface is intentionally a guide rather than a black box: use the code
shown beside each result in an R script or Quarto document to keep the final
analysis reproducible.

The app keeps three table routes separate. Choose **Summary table** for raw
participant-level data, **Epi table** when events and
denominators still need calculation, and **Current data are final calculated
results** in **Customise table** only when every supplied row and value should
be preserved exactly through `as_stats_table()`.

## Five-minute workflow

```{r workflow}
data("birthwt", package = "gtstats")

overview <- describe_data(birthwt)

distribution <- assess_distribution(birthwt, vars = c(age, lwt), by = low)
variance <- assess_variance(birthwt, vars = c(age, lwt), by = low)
# Use `test = "none"` for descriptive spreads only, or `test = "bartlett"`
# when the stronger normal-distribution assumption is justified.

table_one <- summary_table(
  birthwt,
  by = low,
  include = c(age, lwt, race, smoke),
  overall = "last",
  layout = "separate"
) |>
  add_ci(vars = c(age, race)) |>
  add_p()

# Optional categorical-only presentation with distinct n and % columns
categorical_table <- summary_table(
  birthwt,
  by = low,
  include = c(race, smoke),
  categorical_layout = "separate"
)

# Optional compact binary presentation: one clinically relevant event per row
compact_binary_table <- summary_table(
  birthwt,
  by = low,
  include = c(smoke, ht, race),
  show_dichotomous = "single_row",
  value = c(smoke = "Yes", ht = "Yes")
)

# Missing values are excluded from categorical percentage denominators by
# default. Use this explicit option only when Missing is itself a reported
# category and should contribute to 100%.
missing_as_category <- summary_table(
  birthwt,
  include = c(race, smoke),
  missing = "as_category"
)

comparison <- compare_groups(birthwt, variable = lwt, group = low)
comparison$method$selection_rule
diagnostics_stats(comparison)
```

Each object prints as a publication-ready `flextable`, ready for Word and
PowerPoint. Use `customise_table()` to finish its appearance. Use `to_gt()`
only when an HTML-focused `gt` table is specifically required.

```{r export}
table_one |>
  customise_table(
    theme = "journal",
    title = "Table 1. Maternal characteristics",
    spanning_header = list(
      "Birth-weight groups" = c(
        "low = Normal birth weight",
        "low = Low birth weight"
      )
    ),
    borders = "horizontal",
    pvalue_style = "threshold"
  ) |>
  save_output("table-1.docx")

# Explicit HTML/gt route
to_gt(table_one, title = "Table 1. Maternal characteristics")
```

## Browse by task

| Task | Start here |
|---|---|
| First workflow | [Getting started](https://gtstats.thinkdenominator.com/articles/getting-started.html) |
| Complete worked example | [Birth-weight case study](https://gtstats.thinkdenominator.com/articles/birthweight-case-study.html) |
| Build Table 1 | [Descriptive tables](https://gtstats.thinkdenominator.com/articles/descriptive-tables.html) |
| Select and interpret comparisons | [Inferential tests and assumptions](https://gtstats.thinkdenominator.com/articles/inferential-tests.html) |
| Missingness and denominators | [Missing data and denominators](https://gtstats.thinkdenominator.com/articles/missing-data-denominators.html) |
| Confidence intervals and 2x2 tables | [Epidemiology helpers](https://gtstats.thinkdenominator.com/articles/epi-helpers.html) |
| Every argument and default | [Function options](https://gtstats.thinkdenominator.com/articles/function-options.html) |
| All functions | [Reference index](https://gtstats.thinkdenominator.com/reference/index.html) |

## Publication tables and console output

Core statistical functions produce a publication-ready table by default. In a
terminal, automated test, or plain-text workflow, request a real tibble:

```{r console-output, eval=FALSE}
compare_groups(birthwt_data, variable = age, group = low)

compare_groups(
  birthwt_data,
  variable = age,
  group = low,
  format = "tibble"
)
```

The same `format = "table"` or `format = "tibble"` choice is available for
data description, distribution and variance assessment, effect sizes,
correlations, proportions, rates, and cross-tabulations. A summary-table
builder stays composable in either format, so `add_p()` and other layers still
work before the completed tibble is printed.

## Function map

| Workflow | Functions |
|---|---|
| Understand | `describe_data()`, `assess_distribution()`, `assess_variance()` |
| Build Table 1 | `summary_table()`, `add_ci()`, `add_p()`, `add_proportion()`, `add_rate()`, `add_total()`, `add_row()` |
| Compare | `compare_groups()`, `effect_size()`, `correlation()` |
| Outbreak and surveillance tables | `epi_table()` |
| Format an already calculated data frame | `as_stats_table()` |
| Focused epidemiology calculations | `proportion_stats()`, `rate_stats()`, `crosstabs()` |
| Inspect decisions | `assumptions_stats()`, `diagnostics_stats()`, `denominators_stats()` |
| Visualise | `plot_compare()`, `plot_correlation()` |
| Guided interface | `gtstats_app()` |
| Polish and export | `customise_table()`, `to_flextable()`, `to_gt()`, `save_output()` |

`correlation()` handles both a prespecified pair and an exploratory matrix.
The matrix retains pair-specific denominators and p-values in `$summary`, while
`plot_correlation()` provides a labelled coefficient heatmap.

```{r correlation-matrix-readme, eval=FALSE}
correlation(mtcars, vars = c(mpg, disp, hp, wt)) |>
  plot_correlation()
```

## Citation

If `gtstats` contributes to your analysis, tables, teaching, or publication,
please cite:

> Polani R, Kaviprawin M, Sakthivel M, Eliyas SK, Krishnamoorthy Y (2026).
> *gtstats: Beginner-Friendly Statistics and Publication-Ready Tables*.
> R package version 1.0.0. <https://gtstats.thinkdenominator.com/>.

R returns the authoritative citation and BibTeX entry installed with the
package:

```r
citation("gtstats")
toBibtex(citation("gtstats"))
```

The pkgdown **Citing gtstats** article provides a copy-ready BibTeX record and
reproducibility guidance. The reserved Zenodo DOI will be added here after its
record is published and the DOI resolves publicly.
