---
title: "Birth-weight case study: from data to Table 1"
description: "A complete, reproducible gtstats workflow using the birth-weight teaching dataset."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Birth-weight case study: from data to Table 1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# From raw clinical codes to a report-ready table

This case study shows the intended gtstats workflow. A maternity team wants to
describe participants by low-birth-weight outcome, understand the continuous
variables, and produce a Table 1. The example uses the labelled birth-weight
teaching dataset included with gtstats.

## 1. Prepare readable data

```{r prepare}
library(gtstats)
data("birthwt", package = "gtstats")
birthwt_data <- birthwt
```

The built-in data keep the original code names but supply readable variable
labels and factors. `antenatal_visits` is explicitly ordered; numeric clinical
codes are not silently treated as ordinal merely because their values are
ordered.

## 2. Start with the data, not a test

```{r describe}
overview <- describe_data(birthwt_data)
overview
overview$issues
```

`describe_data()` is the first pass. It reports type, completeness, levels or
range, and a short type-specific overview. It does not decide a statistical
test.

## 3. Assess selected continuous variables

```{r distribution}
distribution <- assess_distribution(
  birthwt_data,
  vars = c(age, lwt),
  by = low
)
distribution
distribution$recommendations
```

The first table retains group-specific diagnostics. The recommendation table
gives one reporting approach per variable, so the same summary can be used
across Table 1 groups. Shapiro-Wilk is supporting information, not a rule for
choosing an inferential test. Use `plots = TRUE` when a histogram, density,
Q-Q plot, and boxplot are useful.

`assess_variance()` provides the related spread diagnostic. It reports group
SDs, variances, and ratios without using a variance test to gatekeep Welch
methods.

```{r variance}
variance <- assess_variance(birthwt_data, vars = c(age, lwt), by = low)
variance
```

## 4. Build Table 1

```{r table-one}
table_one <- summary_table(
  birthwt_data,
  by = low,
  include = c(age, lwt, race, smoke, ht, ui,
              previous_preterm, antenatal_visits),
  overall = "first"
) |>
  add_p()

table_one
```

This is already a publication-ready default. You do not need `tbl_stats()` to
see the table. Use it only when you want a `gt` object to modify or export.

```{r render-export}
table_one |>
  tbl_stats(title = "Participant characteristics by birth-weight outcome") |>
  customise_table(theme = "journal") |>
  save_output("table-1.docx")
```

## 5. Ask a focused inferential question

```{r compare}
weight_comparison <- compare_groups(
  birthwt_data,
  variable = lwt,
  group = low
)
weight_comparison

# The exact automatic rule and observed inputs are retained for review.
weight_comparison$method$selection_rule
weight_comparison$method$selection_inputs

assumptions_stats(weight_comparison)
diagnostics_stats(weight_comparison)
denominators_stats(weight_comparison)
```

`compare_groups()` answers one question at a time. The publication table is
the result; the helper functions expose the assumptions, automatic checks, and
analysis denominators for review. Use `effect_size()` separately when the
magnitude of the difference is the main question.

## What this workflow deliberately avoids

- Selecting a test solely from a normality p-value.
- Recommending a different descriptive summary in each group of the same Table 1 variable.
- Treating every numeric code as continuous or ordinal without context.
- Requiring manual formatting before a table can be shared.
