---
title: "Building Descriptive Tables with gtstats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Descriptive Tables with gtstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Introduction

The descriptive table builder is the centrepiece of **gtstats**. It lets you assemble a publication-ready "Table 1" — the standard baseline characteristics table used in clinical and epidemiological papers — through a simple, pipe-based workflow.

The builder has five components:

| Step | Function | What it does |
|---|---|---|
| 1 | `summary_table()` | Initialises the table, sets grouping |
| 2 | `add_summary()` | Adds mean (SD) and median (IQR) rows for continuous variables |
| 3 | `add_proportion()` | Adds n (%) rows for binary/categorical variables |
| 4 | `add_rate()` | Adds event rate rows |
| 5 | `add_total()` | Appends a Total (N) row |
| 6 | `add_p()` | Appends a p-value column |
| 7 | `add_row()` | Appends a free-text row |
| 8 | `tbl_stats()` | Renders the assembled table as a `gt` object |

```{r load}
library(gtstats)
```

---

## A minimal table

Start with `summary_table()`, then pipe in whichever row-builders you need.

```{r minimal}
summary_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  tbl_stats()
```

`by = am` groups the table by the `am` variable (transmission type). Each level becomes a column.

---

## Adding an overall column

Set `overall = TRUE` to include an additional column showing statistics for the full sample.

```{r overall}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  tbl_stats()
```

---

## Summary-table options at a glance

The default call is deliberately simple. These global options cover the most
common reporting choices without requiring separate continuous and categorical
workflows.

| Reporting choice | Option | Example |
|---|---|---|
| Overall column | `overall` | `TRUE`, `"first"`, or `"last"` |
| Continuous summary | `statistic` | `"recommended"`, `"mean_sd"`, `"mean_ci"`, `"median_iqr"`, or `"both"` |
| Categorical display | `categorical` | `"n_percent"`, `"n_over_N_percent"`, `"n"`, or `"percent"` |
| Percentage denominator | `percent` | `"column"`, `"row"`, `"overall"`, or `"none"` |
| Precision | `digits` | `c(continuous = 1, percent = 0, ci = 1)` |
| Missing-value rows | `missing` | `"ifany"`, `"always"`, or `"no"` |
| Categorical confidence intervals | `ci`, `conf.level` | `ci = TRUE, conf.level = 0.95` |
| CI layout | `layout` | `"compact"` (default) or `"separate"` |

For different continuous summaries by variable, provide a named vector:

```{r mixed-continuous-formats}
summary_table(
  mtcars,
  by = am,
  include = c(mpg, wt, cyl),
  statistic = c(mpg = "mean_ci", wt = "median_iqr"),
  overall = "last"
)
```

---

## Choosing percentage denominators and missing rows

Categorical percentages should always identify their denominator:

```{r percentage-denominators}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(cyl, vs), percent = "column")

summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(cyl, vs), percent = "row")

summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(cyl, vs), percent = "overall")

summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(cyl, vs), percent = "none")
```

`"column"` describes levels within each group, `"row"` distributes each level
across groups, `"overall"` uses the full non-missing variable denominator, and
`"none"` displays counts only. The selected rule is retained in the object and
shown in the table footnote.

Missing-value rows are equally explicit:

```{r missing-rows}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(
    vars = c(mpg, wt, vs),
    missing = "ifany"
  )
```

Use `"always"` to show zero-missing rows or `"no"` to suppress them.

### Showing the denominator in every categorical cell

Use `categorical = "n_over_N_percent"` when the table itself should show the
non-missing denominator for every category. This is particularly helpful for
small groups, supplementary tables, and teaching.

```{r n-over-n-percent}
summary_table(
  mtcars,
  by = am,
  include = cyl,
  categorical = "n_over_N_percent",
  percent = "column",
  digits = c(percent = 0)
)
```

### Mean with a confidence interval

Use `statistic = "mean_ci"` when the goal is descriptive precision rather than
a p-value. The interval is a t confidence interval for the observed mean and
uses `conf.level` (95% by default).

```{r mean-ci}
summary_table(
  mtcars,
  include = c(mpg, wt),
  statistic = "mean_ci",
  conf.level = 0.95
)
```

---

## Adding proportion rows

`add_proportion()` adds a highlighted percentage row for a binary or
categorical variable. Optionally add Wilson confidence intervals (the default)
or request exact binomial confidence intervals with
`ci = TRUE`.

When confidence intervals make a table too dense, use
`layout = "separate"`. Each displayed cohort then becomes a spanning header
with a **Summary** (or **n (%)**) column and one **95% CI** column. The compact
layout remains the default for ordinary Table 1 output.

```{r separate-ci-layout}
summary_table(
  birthwt,
  by = smoke,
  include = c(age, low),
  overall = "first",
  ci = TRUE,
  layout = "separate"
)
```

```{r proportion}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  tbl_stats()
```

To display confidence intervals and pin the row to a specific level:

```{r proportion-ci}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs, level = "1", ci = TRUE) |>
  tbl_stats()
```

---

## Adding a total row

`add_total()` appends a row at the bottom showing the total N per group.

```{r total}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  tbl_stats()
```

---

The rendered table footnote is data-driven: it mentions continuous summaries
only when continuous variables are present, and categorical displays only when
categorical variables are present.

## Adding p-values

`add_p()` appends a p-value column. Its automatic route is identical to
`compare_groups(test = "auto")`:

- Continuous variables: Welch t-test (2 groups) or Welch ANOVA (3+ groups)
  by default; use `var_equal = TRUE` only when an equal-variance assumption is
  justified, to select Student's t-test or classical ANOVA
- Marked skewness (absolute sample skewness at least 1 in any group): Wilcoxon
  rank-sum or Kruskal-Wallis. Shapiro-Wilk and lesser asymmetry are supporting
  information and do not switch the test by themselves.
- Binary, nominal, and independent ordinal variables: chi-square when no
  expected count is below 1 and no more than 20% are below 5; Fisher's exact
  test otherwise. Larger sparse tables use a Monte Carlo Fisher p-value.

`add_p()` uses the same automatic-selection policy as `compare_groups()`. The
publication table stays concise: its p-value markers identify the test, while
the variable-specific checks remain available in the audit components.
`var_equal` is a user-specified analytical assumption; it is not inferred by a
variance hypothesis test.

Use `include` when a displayed variable should remain descriptive but should
not be tested. This is important when the grouping variable was derived from a
displayed variable or when a comparison was not prespecified.

```{r add-p-include}
summary_table(birthwt, by = low, include = c(age, bwt, smoke)) |>
  add_p(include = -bwt)
```

For independent ordered factors, Auto compares the complete distribution of
levels using chi-square/Fisher. Specify `method = "wilcox"` or `"kruskal"`
only when the planned estimand is an ordered rank shift. Paired and repeated
outcomes use their design-specific routes: paired t/Wilcoxon signed-rank,
repeated-measures ANOVA/Friedman, McNemar, or Cochran's Q.

```{r add-p}
summary_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_p() |>
  tbl_stats()
```

Raw and multiplicity-adjusted values are retained separately:

```{r adjusted-p}
adjusted <- summary_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt, cyl, vs)) |>
  add_p(p_adjust = "BH")

adjusted$p_values
adjusted$assumptions
adjusted$diagnostics
adjusted$denominators
```

For a readable audit table, use `diagnostics_stats(adjusted)`. For continuous
variables it includes observed group SDs and variances as descriptive context;
these values do not act as a variance-test gatekeeper because Welch methods do
not require equal variances.

`denominators_stats(adjusted)` provides a compact audit table showing the
observations contributing to every variable and group.

### Specifying tests manually

Pass a named character vector to `method` to override the automatic selection for specific variables:

```{r add-p-manual}
summary_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_p(method = c(mpg = "welch_t", wt = "wilcox", cyl = "chisq")) |>
  tbl_stats()
```

Supported methods: `"auto"`, `"welch_t"`, `"t_test"`, `"wilcox"`, `"anova"`,
`"welch_anova"`, `"rm_anova"`, `"kruskal"`, `"friedman"`, `"chisq"`,
`"fisher"`, `"mcnemar"`, and `"cochran_q"`.

### Paired tests

For before/after or matched data, use `paired = TRUE`:

```{r paired}
dat <- data.frame(
  period = c("before", "before", "before", "before",
             "after",  "after",  "after",  "after"),
  score  = c(10, 12, 9, 11, 13, 16, 11, 15)
)

summary_table(dat, by = period) |>
  add_summary(vars = score) |>
  add_p(paired = TRUE, method = "wilcox") |>
  tbl_stats()
```

---

## Adding rate rows

`add_rate()` appends an event rate row calculated per a chosen multiplier, with exact Poisson confidence intervals. This is useful when your dataset contains event counts and person-time denominators.

```{r rate}
summary_table(mtcars, by = am, overall = TRUE, mode = "rate") |>
  add_rate(
    event      = carb,
    time       = cyl,
    label      = "Carburettor rate",
    multiplier = 1000
  ) |>
  tbl_stats()
```

---

## Adding custom rows

`add_row()` inserts a free-text row — useful for study period notes, data source annotations, or any label that does not come from a variable.

```{r custom}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_row(
    label   = "Study period",
    overall = "2020–2024",
    values  = c("am = 1" = "2020–2024", "am = 0" = "2020–2024")
  ) |>
  tbl_stats()
```

---

## The full workflow

Putting it all together — a complete, publication-ready descriptive table:

```{r full-workflow}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs, level = "1", ci = TRUE) |>
  add_total() |>
  add_p() |>
  tbl_stats()
```

---

## Styling the output

`customise_table()` applies a visual theme and relabels columns, rows, and factor levels. It operates on the `gt` object returned by `tbl_stats()`.

```{r style}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p() |>
  tbl_stats() |>
  customise_table(
    theme      = "journal",
    title      = "Table 1. Baseline characteristics by transmission type",
    col_labels = c(
      "Level"  = "",
      "am = 1" = "Manual",
      "am = 0" = "Automatic"
    ),
    row_labels = c(
      "mpg"    = "Miles per gallon",
      "wt"     = "Weight (1000 lbs)",
      "cyl"    = "Cylinders",
      "vs (1)" = "V-shaped engine"
    ),
    accent_color = "#123B7A"
  )
```

Available themes: `"default"`, `"journal"`, `"classic"`, `"minimal"`, `"compact"`.

You can also relabel factor levels within the table using `level_labels`:

```{r level-labels}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = cyl) |>
  tbl_stats() |>
  customise_table(
    level_labels = c(
      "4" = "4-cylinder",
      "6" = "6-cylinder",
      "8" = "8-cylinder"
    )
  )
```

---

## Exporting to Word

Call `to_flextable()` on the `gt_desc_table` object (before rendering with `tbl_stats()`) to produce a `flextable` suitable for Word output.

```{r export}
res <- summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p()

ft <- to_flextable(res)
```

In an R Markdown or Quarto document targeting Word output, simply print `ft` in a chunk and it will appear as a formatted table in the document.

---

## Summary

The gtstats table builder lets you assemble a complete "Table 1" with very little code:

```
summary_table(data, by = group, overall = TRUE)
  |> add_summary(vars = c(...))
  |> add_proportion(var = ...)
  |> add_total()
  |> add_p()
  |> tbl_stats()
  |> customise_table(theme = "journal", ...)
```

Each `add_*()` function is independent — add only the rows your table needs, in any order that makes sense for your report.
