---
title: "Building Descriptive Tables with gtstats"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Descriptive Tables with gtstats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Introduction

The descriptive table builder is the centrepiece of **gtstats**. It lets you assemble a publication-ready "Table 1" — the standard baseline characteristics table used in clinical and epidemiological papers — through a simple, pipe-based workflow.

The builder has five components:

| Step | Function | What it does |
|---|---|---|
| 1 | `descriptive_table()` | Initialises the table, sets grouping |
| 2 | `add_summary()` | Adds mean (SD) and median (IQR) rows for continuous variables |
| 3 | `add_proportion()` | Adds n (%) rows for binary/categorical variables |
| 4 | `add_rate()` | Adds event rate rows |
| 5 | `add_total()` | Appends a Total (N) row |
| 6 | `add_p()` | Appends a p-value column |
| 7 | `add_custom_row()` | Appends a free-text row |
| 8 | `tbl_stats()` | Renders the assembled table as a `gt` object |

```{r load}
library(gtstats)
```

---

## A minimal table

Start with `descriptive_table()`, then pipe in whichever row-builders you need.

```{r minimal}
descriptive_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  tbl_stats()
```

`by = am` groups the table by the `am` variable (transmission type). Each level becomes a column.

---

## Adding an overall column

Set `overall = TRUE` to include an additional column showing statistics for the full sample.

```{r overall}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  tbl_stats()
```

---

## Adding proportion rows

`add_proportion()` adds a row displaying `n (%)` for a binary or categorical variable. Optionally add Wilson confidence intervals with `ci = TRUE`.

```{r proportion}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  tbl_stats()
```

To display confidence intervals and pin the row to a specific level:

```{r proportion-ci}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs, level = "1", ci = TRUE) |>
  tbl_stats()
```

---

## Adding a total row

`add_total()` appends a row at the bottom showing the total N per group.

```{r total}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  tbl_stats()
```

---

## Adding p-values

`add_p()` appends a p-value column. By default, the test is selected automatically:

- Continuous variables: Welch t-test (2 groups) or ANOVA (3+ groups)
- Skewed distributions: Wilcoxon rank-sum or Kruskal-Wallis
- Categorical variables: chi-squared test or Fisher's exact test

```{r add-p}
descriptive_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_p() |>
  tbl_stats()
```

### Specifying tests manually

Pass a named character vector to `method` to override the automatic selection for specific variables:

```{r add-p-manual}
descriptive_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_p(method = c(mpg = "welch_t", wt = "wilcox", cyl = "chisq")) |>
  tbl_stats()
```

Supported methods: `"auto"`, `"welch_t"`, `"t_test"`, `"wilcox"`, `"anova"`, `"kruskal"`, `"chisq"`, `"fisher"`, `"mcnemar"`.

### Paired tests

For before/after or matched data, use `paired = TRUE`:

```{r paired}
dat <- data.frame(
  period = c("before", "before", "before", "before",
             "after",  "after",  "after",  "after"),
  score  = c(10, 12, 9, 11, 13, 16, 11, 15)
)

descriptive_table(dat, by = period) |>
  add_summary(vars = score) |>
  add_p(paired = TRUE, method = "wilcox") |>
  tbl_stats()
```

---

## Adding rate rows

`add_rate()` appends an event rate row calculated per a chosen multiplier, with exact Poisson confidence intervals. This is useful when your dataset contains event counts and person-time denominators.

```{r rate}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_rate(
    events     = carb,
    denom      = cyl,
    label      = "Carburettor rate",
    multiplier = 1000
  ) |>
  tbl_stats()
```

---

## Adding custom rows

`add_custom_row()` inserts a free-text row — useful for study period notes, data source annotations, or any label that does not come from a variable.

```{r custom}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_custom_row(
    label   = "Study period",
    overall = "2020–2024",
    values  = c("am = 1" = "2020–2024", "am = 0" = "2020–2024")
  ) |>
  tbl_stats()
```

---

## The full workflow

Putting it all together — a complete, publication-ready descriptive table:

```{r full-workflow}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs, level = "1", ci = TRUE) |>
  add_total() |>
  add_p() |>
  tbl_stats()
```

---

## Styling the output

`style_table()` applies a visual theme and relabels columns, rows, and factor levels. It operates on the `gt` object returned by `tbl_stats()`.

```{r style}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p() |>
  tbl_stats() |>
  style_table(
    theme      = "journal",
    title      = "Table 1. Baseline characteristics by transmission type",
    col_labels = c(
      "Level"  = "",
      "am = 1" = "Manual",
      "am = 0" = "Automatic"
    ),
    row_labels = c(
      "mpg"    = "Miles per gallon",
      "wt"     = "Weight (1000 lbs)",
      "cyl"    = "Cylinders",
      "vs (1)" = "V-shaped engine"
    ),
    accent_color = "#123B7A"
  )
```

Available themes: `"default"`, `"journal"`, `"classic"`, `"minimal"`, `"compact"`.

You can also relabel factor levels within the table using `level_labels`:

```{r level-labels}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = cyl) |>
  tbl_stats() |>
  style_table(
    level_labels = c(
      "4" = "4-cylinder",
      "6" = "6-cylinder",
      "8" = "8-cylinder"
    )
  )
```

---

## Exporting to Word

Call `to_flextable()` on the `gt_desc_table` object (before rendering with `tbl_stats()`) to produce a `flextable` suitable for Word output.

```{r export}
res <- descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p()

ft <- to_flextable(res)
```

In an R Markdown or Quarto document targeting Word output, simply print `ft` in a chunk and it will appear as a formatted table in the document.

---

## Summary

The gtstats table builder lets you assemble a complete "Table 1" with very little code:

```
descriptive_table(data, by = group, overall = TRUE)
  |> add_summary(vars = c(...))
  |> add_proportion(var = ...)
  |> add_total()
  |> add_p()
  |> tbl_stats()
  |> style_table(theme = "journal", ...)
```

Each `add_*()` function is independent — add only the rows your table needs, in any order that makes sense for your report.
