---
title: "Epidemiology Helpers"
description: >
  Proportions with CIs, event rates with Poisson CIs, and 2x2 tables.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Epidemiology Helpers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE)
```

```{r load}
library(gtstats)
```

## Overview

gtstats provides one table workflow plus three focused epidemiology helpers:

| Function | Purpose |
|---|---|
| `epi_table()` | Multi-outcome outbreak and surveillance tables from line-list or aggregate data |
| `proportion_stats()` | Proportions with Wilson or exact confidence intervals |
| `rate_stats()` | Event rates with exact Poisson confidence intervals |
| `crosstabs()` | Publication-ready categorical cross-tabs; 2×2 tables also give RR, OR and RD |

All functions follow the same pattern: pass in your data frame, name the relevant
variables, and optionally supply a grouping variable with `by =`. They print as
flextables by default; call `to_gt()` for an explicit gt table.

---

## epi_table() — Outbreak and surveillance reporting

`epi_table()` is the table-level workflow. It is the appropriate choice when
several outcomes or strata must be reported together and the numerator,
denominator and uncertainty must remain visible. It does not replace
`summary_table()`: the latter describes participant characteristics, whereas
`epi_table()` estimates occurrence.

### Individual line-list data

```{r epi-line}
data("birthwt", package = "gtstats")
epi_line <- epi_table(
  birthwt,
  outcomes = low,
  by = smoke,
  event = "Low birth weight",
  measure = "prevalence",
  p_value = TRUE,
  effects = "all"
)
to_flextable(epi_line)
```

For each outcome, the denominator is the number of non-missing records in the
relevant group. A case-only line list therefore cannot estimate an attack rate;
use aggregate data with an external population denominator.

### Aggregate surveillance data

```{r epi-aggregate}
surveillance <- data.frame(
  ward = c("A", "B", "C"),
  cases = c(12, 7, 4),
  population = c(80, 65, 52)
)
epi_aggregate <- epi_table(
  surveillance,
  numerator = cases,
  denominator = population,
  by = ward,
  label = "Influenza",
  measure = "attack_rate",
  multiplier = 100
)
to_flextable(epi_aggregate)
```

For incidence rates, the denominator is accumulated person-time and the
interval is exact Poisson. Optional p-values compare groups. Effect estimates
are available only for exactly two groups, because their direction must be
unambiguous. Inspect `$denominators`, `$p_values` and `$effects` before
reporting.

---

## proportion_stats() — Proportions

`proportion_stats()` calculates the proportion of a selected level together with a Wilson score confidence interval by default. Exact binomial intervals remain available with `ci_method = "exact"`.

Its publication table keeps the event count and percentage together as
`n (%)` and places the confidence interval in a separate column. When `by` is
used, each group becomes a spanning header above these two columns. The tidy
long-form numerical results remain available from `$summary`.

```{r prop-basic}
# Overall proportion
to_flextable(proportion_stats(mtcars, var = vs))
```

```{r prop-grouped}
# Grouped by transmission type
to_flextable(proportion_stats(mtcars, var = vs, by = am))
```

```{r prop-level}
# Pin to a specific level
to_flextable(proportion_stats(mtcars, var = vs, by = am, level = "1"))
```

```{r prop-gt}
to_gt(proportion_stats(mtcars, var = vs, by = am))
```

### In a descriptive table

`add_proportion()` embeds the same calculation inside the table builder:

```{r prop-table}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_proportion(var = vs, level = "1", ci = TRUE) |>
  add_total() |>
  to_gt()
```

---

## rate_stats() — Event rates

`rate_stats()` divides the total number of events by the total person-time and multiplies by a chosen denominator (e.g. 1 000 person-years). Confidence intervals are calculated using the exact Poisson method.

The publication table places each group above four separate columns: events,
accumulated person-time, rate and confidence interval. This keeps the numerator,
denominator and uncertainty visible without repeating interval labels inside
every rate cell. Record counts and all numerical values remain in `$summary`.

```{r rate-data}
df <- data.frame(
  event = c(1, 0, 1, 0, 1, 1),
  ptime = c(10, 12, 8, 9, 11, 7),
  arm   = c("A", "A", "A", "B", "B", "B")
)
```

```{r rate-basic}
# Overall rate
to_flextable(rate_stats(df, event = event, time = ptime))
```

```{r rate-grouped}
# Grouped by study arm
to_flextable(rate_stats(df, event = event, time = ptime, by = arm))
```

```{r rate-gt}
rate_stats(df, event = event, time = ptime, by = arm) |>
  to_gt()
```

### In a descriptive table

`add_rate()` embeds rate rows in the table builder:

```{r rate-table}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_rate(
    event      = carb,
    time       = cyl,
    label      = "Carburettor rate (per 1 000)",
    multiplier = 1000
  ) |>
  to_gt()
```

---

## crosstabs() — categorical tables and 2×2 epidemiology

`crosstabs()` takes any two categorical variables. When both are binary, `row`
defines the exposure/reference axis and `col` the outcome/event axis, so risks
and ratios retain an explicit direction. The 2×2 output additionally calculates:

- Risk in the exposed and unexposed groups, with Wilson 95% CIs
- Risk ratio (RR) with 95% CI
- Risk difference with a Newcombe hybrid-score CI
- An automatically selected association test: Fisher's exact test when
  expected counts are sparse, otherwise the chi-squared test

It also supports any categorical table size. By default cells show `n` and
column percentages, with row/column/grand totals. Use `percent = "row"` for
within-row outcome frequencies, or `percent = c("row", "column")` when both
denominators are useful in an exploratory or supplementary table.

```{r crosstabs-multilevel}
to_flextable(crosstabs(mtcars, row = cyl, col = am))
to_flextable(crosstabs(mtcars, row = cyl, col = gear, percent = c("row", "column")))
```

For tables larger than 2×2, the footer reports the association test and
Cramér's V. Risk ratio, odds ratio, and risk difference are intentionally only
available for a binary 2×2 contrast.

The odds ratio is available when it is the effect measure the study needs,
but is deliberately not part of the routine default.

```{r twobytwo}
to_flextable(crosstabs(mtcars, row = am, col = vs))
```

```{r twobytwo-gt}
to_gt(crosstabs(mtcars, row = am, col = vs))
```

Choose the direction explicitly when the coding or factor order does not
express the scientific question:

```{r twobytwo-direction}
to_flextable(crosstabs(
  mtcars,
  row = am,
  col = vs,
  row_level = 1,
  col_level = 1
))
```

Whether levels are supplied or selected automatically, inspect
`result$inputs$row_level` and `result$inputs$col_level` before reporting a
2×2 measure. The audit retains the selected exposure/event direction,
expected-count rule, zero-cell strategy, and complete-pair denominators while
the publication table stays concise.

For a case-control analysis, request the odds ratio. The association test can
also be omitted when the table is intended to report effects and confidence
intervals only:

```{r twobytwo-or}
to_flextable(crosstabs(
  mtcars,
  row = am,
  col = vs,
  measures = "or",
  test = "none"
))
```

When a cell is zero, log-scale ratio intervals use the Haldane–Anscombe
correction by default. The strategy is explicit and configurable:

```{r zero-cell}
to_flextable(crosstabs(
  mtcars,
  row = am,
  col = vs,
  zero_correction = "none"
))
```

Without correction, an infinite or zero ratio may be reported and its
log-scale confidence interval may be unavailable.

---

## Tips

- Use `risk_ci = "exact"` when an exact binomial interval is required by the
  reporting convention; Wilson is the routine default.
- `test = "auto"` checks expected counts and chooses Fisher's exact test for a
  sparse table.
- `proportion_stats()`, `rate_stats()`, and `crosstabs()` all return structured
  objects that print as flextables, can be rendered with `to_gt()`, or can be
  inspected as plain lists.
- Inside a descriptive table, use `add_proportion()` and `add_rate()` rather than standalone helpers to keep everything in one table.
