---
title: "Epidemiology Helpers"
description: >
  Proportions with CIs, event rates with Poisson CIs, and 2x2 tables.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Epidemiology Helpers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r load}
library(gtstats)
```

## Overview

gtstats includes three standalone helper functions aimed specifically at clinical and epidemiological analysis:

| Function | Purpose |
|---|---|
| `prop_ci()` | Proportions with Wilson confidence intervals |
| `rate_stats()` | Event rates with exact Poisson confidence intervals |
| `twobytwo_table()` | 2×2 table with risk ratio, odds ratio, and CIs |

All three follow the same pattern: pass in your data frame, name the relevant variables, and optionally supply a grouping variable with `by =`. Pipe to `tbl_stats()` for a formatted `gt` table.

---

## prop_ci() — Proportions

`prop_ci()` calculates the proportion of a binary variable that equals a specified level (default: the first non-NA value), together with a Wilson confidence interval.

```{r prop-basic}
# Overall proportion
prop_ci(mtcars, var = vs)
```

```{r prop-grouped}
# Grouped by transmission type
prop_ci(mtcars, var = vs, by = am)
```

```{r prop-level}
# Pin to a specific level
prop_ci(mtcars, var = vs, by = am, level = "1")
```

```{r prop-gt}
tbl_stats(prop_ci(mtcars, var = vs, by = am))
```

### In a descriptive table

`add_proportion()` embeds the same calculation inside the table builder:

```{r prop-table}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_proportion(var = vs, level = "1", ci = TRUE) |>
  add_total() |>
  tbl_stats()
```

---

## rate_stats() — Event rates

`rate_stats()` divides the total number of events by the total person-time and multiplies by a chosen denominator (e.g. 1 000 person-years). Confidence intervals are calculated using the exact Poisson method.

```{r rate-data}
df <- data.frame(
  event = c(1, 0, 1, 0, 1, 1),
  ptime = c(10, 12, 8, 9, 11, 7),
  arm   = c("A", "A", "A", "B", "B", "B")
)
```

```{r rate-basic}
# Overall rate
rate_stats(df, event = event, time = ptime)
```

```{r rate-grouped}
# Grouped by study arm
rate_stats(df, event = event, time = ptime, by = arm)
```

```{r rate-gt}
rate_stats(df, event = event, time = ptime, by = arm) |>
  tbl_stats()
```

### In a descriptive table

`add_rate()` embeds rate rows in the table builder:

```{r rate-table}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_rate(
    events     = carb,
    denom      = cyl,
    label      = "Carburettor rate (per 1 000)",
    multiplier = 1000
  ) |>
  tbl_stats()
```

---

## twobytwo_table() — 2×2 epi tables

`twobytwo_table()` takes an exposure variable and a binary outcome, constructs the 2×2 table, and calculates:

- Cell counts (a, b, c, d)
- Risk ratio (RR) with 95% CI
- Odds ratio (OR) with 95% CI
- Chi-squared p-value

```{r twobytwo}
twobytwo_table(mtcars, exposure = am, outcome = vs)
```

```{r twobytwo-gt}
tbl_stats(twobytwo_table(mtcars, exposure = am, outcome = vs))
```

---

## Tips

- For small cell counts (fewer than 5), consider reporting Fisher's exact p-value from `compare_groups()` alongside the 2×2 table.
- `prop_ci()`, `rate_stats()`, and `twobytwo_table()` all return structured objects that can be rendered with `tbl_stats()` or inspected as plain lists.
- Inside a descriptive table, use `add_proportion()` and `add_rate()` rather than standalone helpers to keep everything in one table.
