
---
title: "Getting Started"
description: >
  Learn the gtstats workflow: explore, compare, build Table 1, and export results.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

### Installation
```{r install}
# remotes::install_github("thinkdenominator/gtstats")
install.packages("gtstats")
library(gtstats)
```


⸻

The gtstats philosophy

gtstats is designed around a simple workflow:

	•	User-friendly API and args.
	•	Tidy outputs
	•	Build a Table 1 and test hypotheses with minimal code
	•	Export ready-to-use outputs

⸻

## Explore your data

Start by understanding variable types and missingness.

```{r describe}
describe_data(mtcars)
# tidy output
describe_data(mtcars) |> tbl_stats() 
```

⸻

## Summary statistics

Get quick descriptive summaries:
```{r summary}
summary_stats(mtcars, by = am)

Render as a table:

summary_stats(mtcars, by = am) |> tbl_stats()
```


⸻

## Compare groups

Compare outcomes across groups with automatic test selection:

```{r compare}
compare_groups(mtcars, outcome = mpg, group = am) |>
  tbl_stats()
```
👉 gtstats chooses the test for you
👉 Output tells you which test was used

⸻

## Visualise comparisons

Plots are essential for interpretation.

```{r plot}
plot_compare(
  mtcars,
  outcome = mpg,
  group = am,
  show_p = TRUE
)
```


	•	Continuous → boxplot
	•	Categorical → bar chart
	•	Optional p-value overlay

⸻

## Build Table 1 

This is the heart of gtstats.

```{r table1}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl))  |>
  add_proportion(var = vs, ci = TRUE)  |>
  add_total()                          |>
  add_p()                              |>
  tbl_stats()
```

✔ Continuous: mean (SD)
✔ Categorical: n (%)
✔ Proportions: with CI
✔ Tests: automatically applied

⸻

## Style outputs

```{r style}
tbl_stats(
  descriptive_table(mtcars, by = am) |>
    add_summary(vars = c(mpg, wt)) |>
    add_p()
) |>
  style_table(
    theme = "journal",
    accent_color = "#1C1C1E"
  )
```

⸻

## Export results


### Save tables
```{r export}
res <- descriptive_table(mtcars, by = am) |>
  add_summary(vars = c(mpg, wt)) |>
  add_p()

save_table(res, "table1.html")
save_table(tbl_stats(res), "table1.png")

```
⸻

### Save plots
```{r export-plot}
p <- plot_compare(mtcars, mpg, am, show_p = TRUE)

save_plot(p, "mpg_by_am.png")
```
You can also control resolution:
```{r export-plot-res}
save_plot(p, "figure.png", width = 7, height = 5, dpi = 300)
```

⸻

## Epidemiology helpers

```{r epi}
prop_ci(mtcars, var = vs, by = am)

rate_stats(df, event = event, time = ptime, by = arm)

twobytwo_table(mtcars, exposure = am, outcome = vs)
```

⸻

