---
title: "Start here: a simple gtstats workflow"
description: "Understand data, make a publication-ready descriptive table, and compare groups."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Start here: a simple gtstats workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# The workflow

gtstats is designed around one practical sequence:

1. **Understand** the data with `describe_data()`.
2. **Assess** selected continuous variables with `assess_distribution()` and,
   for grouped data, `assess_variance()`.
3. **Describe** participants with `summary_table()`.
4. **Compare** one variable across groups with `compare_groups()`.
5. **Inspect** assumptions, diagnostics, and denominators when needed.
6. **Render or export** only when you want to customise the finished output.

The main functions print publication-ready tables by default. `tbl_stats()` is
not an analysis step: it provides the `gt` table for titles, styling, and
export.

## Install and load

```{r install}
# Development version until the package is on CRAN
remotes::install_github("ThinkDenominator/gtstats")
library(gtstats)
data("birthwt", package = "gtstats")
```

## 1. Understand a dataset

```{r first-look}
overview <- describe_data(birthwt)
overview
overview$issues
```

The printed table is a compact first look. `$issues` is deliberately separate
so an ordinary data overview is not cluttered with cautions that do not apply.

## Built-in teaching datasets

Use the labelled datasets included with gtstats when learning the workflow or
sharing reproducible examples:

```r
data("birthwt", "trial_data", "paired_data", package = "gtstats")
```

`birthwt` is a real low-birth-weight study dataset. `trial_data` is a
three-arm synthetic clinical trial designed to demonstrate Welch ANOVA,
Kruskal-Wallis, chi-square, Fisher exact, correlation, and rates.
`paired_data` is long-format paired follow-up data for paired t-tests,
Wilcoxon signed-rank, and McNemar tests. Each help page contains its data
dictionary and examples.

## Prefer a guided interface?

`gtstats_app()` is an optional Shiny companion for learners and collaborators
who prefer a menu-driven starting point. It can load a built-in teaching
dataset or a CSV/Excel file, show a working data dictionary, then guide the
user through data description, distribution and spread checks, Table 1, group
comparisons, and crosstabs. Every analysis page displays the corresponding R
code with copy and `.R` download buttons, while result tables can be downloaded
as Word, HTML, PDF, or RTF. The optional `rio` package enables Excel import.

```r
# Run once if Shiny is not installed
install.packages("shiny")
install.packages("rio") # only needed for Excel files

gtstats_app()
```

The GUI is not a substitute for a reproducible script: copy the displayed code
into an R script or Quarto document before final reporting. While the app is
open, the R console displays `Listening on ...`; this is normal. Click **Close
app** in the bottom-right corner of the interface to end the local session
cleanly and return to the R prompt.

## 2. Assess distribution for descriptive reporting

```{r assess}
distribution <- assess_distribution(birthwt, vars = c(age, lwt), by = low)
distribution
distribution$recommendations
```

This function is restricted to continuous numeric variables. It uses
skewness, data quality, and Shapiro-Wilk as supporting information to guide
descriptive presentation. It does not choose an inferential test.

When groups are being compared, inspect their observed spread separately:

```{r assess-variance}
variance <- assess_variance(birthwt, vars = c(age, lwt), by = low)
variance
```

`assess_variance()` reports SDs, variances, and largest/smallest ratios. These
are descriptive diagnostics, not a test-selection rule: Welch methods do not
require equal variances.

## 3. Create a Table 1

```{r table-one}
table_one <- summary_table(
  birthwt,
  by = low,
  include = c(age, lwt, race, smoke),
  overall = "last"
) |>
  add_p()

table_one
```

Add optional rows only when they answer a specific reporting need:

```{r optional-rows}
table_one |>
  add_proportion(var = smoke, level = "Yes", label = "Smoking prevalence", ci = TRUE) |>
  add_total()
```

## 4. Compare one variable

```{r compare}
comparison <- compare_groups(birthwt, variable = lwt, group = low)
comparison
```

For a transparent audit trail:

```{r audit}
assumptions_stats(comparison)
diagnostics_stats(comparison)
denominators_stats(comparison)
```

## 5. Customise and export

```{r export}
table_one |>
  tbl_stats(title = "Table 1. Vehicle characteristics") |>
  customise_table(theme = "journal") |>
  save_output("table-1.docx")
```

Before reporting any percentage or inference, read [Missing data and denominators](missing-data-denominators.html). For a full clinical example, continue to the [birth-weight case study](birthweight-case-study.html).
