---
title: "Missing data and denominators"
description: "How gtstats handles missing values, percentages, complete cases, and analysis denominators."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing data and denominators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE)
```

```{r load}
library(gtstats)
```

## The policy in one minute

gtstats does not silently treat missing values as observations with a value.
Each analysis retains or reports the observations contributing to its result.
The usual default is to use observations with the values needed for that
calculation and make the resulting denominator available for review.

There is no universal denominator. A categorical percentage needs observed
values of that variable; a correlation needs complete pairs; a rate needs valid
event-time records. The right denominator depends on the scientific question.

1. Start with `describe_data()` to see completeness for every variable.
2. Choose the percentage denominator deliberately for descriptive tables.
3. Use `denominators_stats()` before reporting results when exclusions matter.
4. Describe the analysis population when missing data are material.

## Key terms

| Term | Meaning in gtstats |
|---|---|
| `n_total` | Rows in the relevant dataset or group before the statistic-specific exclusion |
| `n_nonmissing` | Rows with the values required for the variable |
| `n_missing` | Rows excluded because a required value is missing |
| `level` | Categorical level represented by a displayed cell, when relevant |
| `numerator` | Count of events or selected levels, where relevant |
| `denominator` | Eligible observations or total person-time used for the estimate |
| `rule` | Plain-language record of how the denominator was formed |

`denominators_stats()` returns a formatted table by default, or a plain console
tibble with `format = "tibble"`. For categorical summaries, it records the level
as well as the group, so row and overall percentages can be checked against the
exact denominator used in the displayed cell.

## Start by describing missingness

`describe_data()` reports completeness from all rows in the supplied dataset;
it never drops rows to make the overview look complete.

```{r data}
dat <- data.frame(
  arm = factor(c("Control", "Control", "Treatment", "Treatment", "Treatment")),
  age = c(45, NA, 51, 62, 57),
  smoker = factor(c("No", "Yes", NA, "Yes", "No")),
  follow_up = c(12, 10, 8, NA, 11)
)

to_flextable(describe_data(dat))
```

This reports the amount of missing data, not its cause. Clinical context, the
data dictionary, and the study design are needed to decide whether a
complete-case analysis is scientifically appropriate.

## Descriptive tables: percentage denominators

For categorical variables, `summary_table()` makes the
denominator explicit. The default is `percent = "column"`.

| Setting | Percentage denominator | Typical use |
|---|---|---|
| `"column"` | Non-missing observations in each displayed group | Standard Table 1 |
| `"row"` | Non-missing observations for that level across groups | Distribution of a category across groups |
| `"overall"` | All non-missing observations of the variable | Share of the overall sample |
| `categorical = "n"` | No percentage | Counts only |

```{r descriptive}
summary_table(
  dat,
  by = arm,
  include = smoker,
  percent = "column",
  missing = "ifany"
) |> to_flextable()

to_flextable(summary_table(dat, by = arm, include = smoker, percent = "row"))
to_flextable(summary_table(dat, by = arm, include = smoker, percent = "overall"))
to_flextable(summary_table(dat, by = arm, include = smoker, categorical = "n"))
```

`missing = "ifany"` displays a Missing row only when it is needed. Use
`"always"` to show zero-missing rows or `"no"` to suppress them. Suppressing a
row does not put missing values back into a categorical denominator.

Continuous summaries use observed finite values of that variable. A displayed
Missing row is calculated against all rows in the relevant group.

Use `categorical = "n_over_N_percent"` when the table should show the
non-missing denominator inside every cell, for example `30/59 (50.8%)`.
For a continuous descriptive estimate with precision but no p-value, use
`statistic = "mean_ci"`; this reports a t confidence interval for the observed
mean using `conf.level`.

## Proportions and confidence intervals

`proportion_stats()` and `add_proportion()` use non-missing observations of the
selected variable as the denominator. The numerator is the chosen level, and
the confidence interval describes that observed-data proportion.

```{r proportion}
smoking <- proportion_stats(dat, smoker, by = arm, level = "Yes")
to_flextable(smoking)
denominators_stats(smoking)
```

If unknown status should be included in the clinical denominator, do not assume
the observed-data proportion answers that question. Report unknown values or
define the required analysis approach in advance.

An all-missing categorical variable remains visible in `summary_table()` as an
em dash, with a Missing row by default. It is not silently dropped from the
table.

## Rates use complete event-time records

`rate_stats()` and `add_rate()` require finite, non-negative event counts and
person-time values. Rows missing either value are excluded. The denominator is
the sum of valid person-time, not the number of patients. A group with zero
accumulated time is retained but its rate is reported as not estimable.

```{r rate}
rate_dat <- data.frame(
  arm = c("Control", "Control", "Treatment", "Treatment"),
  events = c(1, 0, 2, 1),
  person_years = c(1.2, NA, 0.8, 1.0)
)

rate <- rate_stats(rate_dat, event = events, time = person_years, by = arm)
denominators_stats(rate)
```

The audit records both valid event-time records and the person-time denominator.

## Cross-tabs, comparisons, and correlations

`crosstabs()` uses complete row-column pairs: a row is excluded when either
cross-tabulated variable is missing. Its margins, percentages, test, and 2×2
measures therefore refer to that same complete-pair population.

```{r crosstab}
cross <- crosstabs(dat, row = arm, col = smoker)
denominators_stats(cross)
```

For independent `compare_groups()`, an observation requires both `variable` and
`group`. For paired analysis, an identifier must be observed once in each group
with the required outcome; the denominator becomes the number of complete,
aligned pairs.

```{r comparison}
comparison <- compare_groups(dat, variable = age, group = arm)
denominators_stats(comparison)
```

`correlation()` similarly uses complete pairs: both variables must be observed
in the same row.

```{r correlation}
cor_dat <- data.frame(
  age = c(34, 41, 45, 49, 53, 57, 62, 68),
  follow_up = c(12, 11, NA, 9, 8, 7, 6, 5)
)
to_flextable(correlation(cor_dat, x = age, y = follow_up))
```

## Distribution assessment

`assess_distribution()` keeps data-quality information visible. It reports
`n`, `Missing`, and `Non-finite`; skewness and Shapiro-Wilk use usable finite
values only. Missing or non-finite values are flagged in the guidance rather
than being interpreted as evidence of non-normality.

```{r distribution}
dist_dat <- data.frame(
  arm = factor(rep(c("Control", "Treatment"), each = 8)),
  age = c(34, 39, 44, 48, 52, 57, NA, 63,
          36, 41, 46, 51, 56, 61, 66, Inf)
)
to_flextable(assess_distribution(dist_dat, vars = age, by = arm))
```

For a grouped continuous variable, `assess_variance()` also excludes missing
and non-finite values from its SD and variance calculations, while retaining
their counts in the output.

```{r variance}
to_flextable(assess_variance(dist_dat, vars = age, by = arm))
```

## Reporting checklist

- Does the displayed percentage use the denominator that answers the question?
- Are missing values shown, suppressed, or described elsewhere?
- Is the analysis population the same for every variable? If not, is this clear?
- For rates, is the denominator person-time rather than participants?
- For paired analysis, have you reported the number of complete pairs?
- For repeated-measures ANOVA, have you reviewed sphericity rather than using a
  cross-occasion variance test as a substitute?
- Have you reviewed `denominators_stats()` where exclusions affect interpretation?

## What gtstats does not do automatically

gtstats reports the observed-data calculation. It does not impute missing
values, diagnose the missing-data mechanism, apply weighting, or decide whether
complete-case analysis is scientifically valid. Prepare the appropriate analysis
dataset first when those methods are required, then use gtstats to report the
result transparently.
