---
title: "Package Overview"
description: >
  A quick tour of every function in gtstats and how they fit together.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r load}
library(gtstats)
```

## Why gtstats?

Most descriptive statistics tasks in clinical and public health work boil down to the same handful of steps: summarise your variables, check whether they are normally distributed, compare groups, test associations, and build a clean table for your report or paper. **gtstats** provides a single, consistent interface for all of these steps, with output that goes straight into a `gt` table or a Word document.

---

## Function map

```
Data exploration ─────── describe_data()

Summary statistics ───── summary_stats()
Distribution checks ──── check_distribution()
Group comparisons ────── compare_groups()
Correlations ─────────── correlate_vars()

Proportions + CI ──────── prop_ci()
Rates + Poisson CI ────── rate_stats()
2×2 epi tables ─────────  twobytwo_table()

Table builder ──────────  descriptive_table()
                          ├─ add_summary()
                          ├─ add_proportion()
                          ├─ add_rate()
                          ├─ add_total()
                          ├─ add_p()
                          └─ add_custom_row()

Render ─────────────────  tbl_stats()
Style ──────────────────  style_table()
Export ─────────────────  to_flextable()
```

---

## Typical usage patterns

### Quick data check

```{r}
describe_data(mtcars)
summary_stats(mtcars, by = am)
```

### One-step group comparison

```{r}
compare_groups(mtcars, outcome = mpg, group = am) |>
  tbl_stats()
```

### Epidemiology helpers

```{r}
prop_ci(mtcars, var = vs, by = am)

twobytwo_table(mtcars, exposure = am, outcome = vs) |>
  tbl_stats()
```

### Full descriptive table

```{r}
descriptive_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p() |>
  tbl_stats() |>
  style_table(theme = "journal", accent_color = "#123B7A")
```

---

## Output formats

Every analysis function returns a structured list with a class attribute. Printing it in the console gives a plain-text summary. Passing it to `tbl_stats()` renders a formatted `gt` table. The descriptive table builder also supports `to_flextable()` for Word-ready output.

---

## Getting help

- See the [Getting Started](getting-started.html) article for a step-by-step introduction.
- See the [Descriptive Tables](descriptive-tables.html) article for a deep dive on the table builder.
- Browse the [Reference](../reference/index.html) for full argument documentation.
