---
title: "Function map"
description: "Choose the gtstats function that fits your analytical task."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Function map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# Choose by task

| Question | Start with | Use next when needed |
|---|---|---|
| What is in my dataset? | `describe_data()` | `assess_distribution()` and `assess_variance()` for selected continuous variables |
| How do I describe participants by group? | `summary_table()` | `add_p()`, `add_total()`, or `add_proportion()` only when justified |
| Do groups differ for one variable? | `compare_groups()` | `effect_size()` for magnitude; audit helpers for review |
| Are two continuous variables associated? | `correlation()` | `plot_correlation()` |
| What is a proportion or rate? | `proportion_stats()` or `rate_stats()` | — |
| What are RR, OR, and risk difference? | `crosstabs()` | — |
| How do I change the finished table? | `tbl_stats()` | `customise_table()`, `to_flextable()`, `save_output()` |

## The core distinction

`summary_table()` describes **many variables** for a descriptive table.
`compare_groups()` tests **one variable** across one group variable.
`assess_distribution()` describes the empirical shape of selected continuous
variables, not an inferential test choice.

`assess_variance()` describes group SDs, variances, and spread ratios for
selected continuous variables. It is diagnostic only: Welch methods do not
require equal variances, and the function does not select an inferential test.

## Objects remain inspectable

Every main function returns an object containing the data behind the display.
The default print method is the publication-ready table. Components such as
`$summary`, `$inferential`, `$recommendations`, and `$notes` are available for
programming and review; they are not extra steps required for routine use.

```{r example}
result <- compare_groups(mtcars, variable = mpg, group = am)
result$inferential
diagnostics_stats(result)
```

## A minimal decision tree

```text
Raw dataset
  └─ describe_data()
      ├─ assess_distribution()  → how to present selected continuous variables
      ├─ assess_variance()      → how observed spread differs across groups
      ├─ summary_table()        → Table 1 / participant characteristics
      └─ compare_groups()       → one inferential question
```

The [Start here](getting-started.html) article explains the minimal workflow;
the [birth-weight case study](birthweight-case-study.html) applies it end to
end. [Missing data and denominators](missing-data-denominators.html) explains
how every analysis population and displayed percentage is formed.

For a compact, user-facing table of arguments and defaults for every exported
function, see [Function options](function-options.html).
