---
title: "Styling and Exporting Tables"
description: >
  Finish publication tables, choose an output engine, and export to Office or HTML.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Styling and Exporting Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE)
```

```{r load}
library(gtstats)
```

## The output model

gtstats separates analysis from presentation:

1. Build the statistical result with `summary_table()` and optional layers.
2. Finish its appearance with `customise_table()`.
3. Export the same object with `save_output()`.

Publication results print as `flextable` objects by default because they travel
cleanly to Word and PowerPoint. `to_gt()` is the explicit HTML-focused route.
`to_flextable()` remains available when an explicit conversion is useful.

```{r result}
res <- summary_table(
  mtcars,
  by = am,
  include = c(mpg, wt, cyl),
  overall = TRUE,
  statistic = c(continuous = "mean_sd", wt = "median_iqr")
) |>
  add_ci(vars = c(mpg, cyl)) |>
  add_p()

res                   # default flextable
to_gt(res)            # explicit gt table
to_flextable(res)     # explicit flextable
```

The reserved name `continuous` supplies a default for all selected continuous
variables. A named variable then overrides it. Variables not covered by a
global or specific rule retain the package recommendation.

## Format statistics calculated elsewhere

If a data frame already contains the final numbers, do not pass it to
`summary_table()`: that would summarise those numbers again. Wrap it with
`as_stats_table()` instead. The wrapper preserves the supplied rows and values
and makes the object compatible with the same styling and export tools.

```{r calculated-table}
calculated <- mtcars |>
  dplyr::group_by(am) |>
  dplyr::summarise(
    N = dplyr::n(),
    `Mean mpg` = round(mean(mpg), 1),
    .groups = "drop"
  )

calculated_table <- as_stats_table(
  calculated,
  notes = "Values were calculated before table formatting."
)

customise_table(calculated_table, title = "Summary supplied by the analyst")
```

If the supplied data contain the ingredients for a confidence interval, map
them explicitly before styling. For example, event counts and person-time can
be converted to exact Poisson rate intervals:

```{r aggregate-rate-ci}
rates <- data.frame(
  Group = c("Intervention", "Control"),
  Events = c(8, 14),
  PersonYears = c(420, 510)
)

rate_table <- as_stats_table(rates) |>
  add_ci(
    type = "rate",
    numerator = Events,
    denominator = PersonYears,
    multiplier = 1000
  )

customise_table(rate_table, title = "Events per 1,000 person-years")
```

The mapping is intentionally explicit: gtstats does not guess denominators or
statistical meaning from uploaded column names.

The distinction is important:

| What each row represents | Correct starting function |
|---|---|
| Participant or ordinary observation | `summary_table()` |
| Outbreak/surveillance observation or aggregate numerator and denominator | `epi_table()` |
| Final result calculated upstream | `as_stats_table()` |

An `as_stats_table()` result prints as a publication-ready flextable and can
be passed directly to `customise_table()`, `to_flextable()`, `to_gt()`, or
`save_output()`. Styling never changes its supplied values.

## Themes and journal styling

```{r themes}
customise_table(res, theme = "journal")
customise_table(res, theme = "minimal")
customise_table(res, theme = "compact")
```

| Theme | Typical use |
|---|---|
| `"default"` | General reports |
| `"journal"` | Manuscripts and appendices |
| `"classic"` | Traditional ruled tables |
| `"minimal"` | Slides and modern reports |
| `"compact"` | Dense tables |

`borders` independently controls the rule pattern, while `density` changes
cell padding:

```{r borders-density}
customise_table(
  res,
  theme = "journal",
  borders = "horizontal",
  density = "compact",
  font = "Arial",
  font_size = 10
)
```

## Titles, spanning headers and footnotes

```{r headers-notes}
customise_table(
  res,
  title = "Table 1. Vehicle characteristics",
  subtitle = "Grouped by transmission",
  spanning_header = list(
    "Transmission groups" = c("am = 0", "am = 1")
  ),
  footnotes = c(
    "CI = confidence interval.",
    "P-values are two-sided."
  ),
  show_footnotes = TRUE
)
```

`show_footnotes = FALSE` removes the automatically generated analytical
footnotes. `footnotes` adds only the user-supplied notes. This changes the
presentation, not the underlying analysis.

## Relabelling and column control

Named vectors use the completed output value as the name and the desired
display text as the value.

```{r relabel}
customise_table(
  res,
  col_labels = c(
    "am = 0" = "Automatic",
    "am = 1" = "Manual",
    "Overall" = "All vehicles"
  ),
  row_labels = c(
    "mpg" = "Fuel economy",
    "wt" = "Weight",
    "cyl" = "Cylinders"
  ),
  level_labels = c(
    "4" = "Four",
    "6" = "Six",
    "8" = "Eight"
  ),
  hide_cols = "Level",
  column_widths = c(Variable = 2.2)
)
```

Use `align`, `bold_cols`, and `italic_cols` for targeted emphasis. These
arguments accept completed column names.

## P-value display

P-value styling changes only how already calculated p-values are printed.

```{r pvalue}
customise_table(
  res,
  pvalue_style = "threshold",
  pvalue_digits = 3,
  pvalue_threshold = 0.001,
  pvalue_prefix = TRUE
)
```

Available styles are `"threshold"`, `"fixed"`, and `"scientific"`.

## Choosing the renderer

`customise_table()` defaults to `engine = "flextable"`:

```{r engines}
office_table <- customise_table(res, theme = "journal")
html_table <- customise_table(res, theme = "journal", engine = "gt")
```

Publication footnotes use smaller type than the body in both engines (8 pt in
flextable and 10 px in gt), matching the gtregression house style. They remain
readable in Word while taking less vertical space in long tables.

Use flextable for Word, PowerPoint, RTF, and Office-centred workflows. Use gt
for HTML pages or when downstream code specifically expects a `gt_tbl`.

## Export

`save_output()` accepts the original gtstats result, a customised flextable, a
gt table, or a ggplot.

```{r save, eval=FALSE}
finished <- customise_table(
  res,
  theme = "journal",
  title = "Table 1. Vehicle characteristics"
)

save_output(finished, "table-1.docx")
save_output(finished, "table-1.pptx")
save_output(finished, "table-1.rtf")

html_table <- customise_table(res, engine = "gt")
save_output(html_table, "table-1.html")
```

Table extensions include `.docx`, `.pptx`, `.rtf`, `.html`, `.png`, `.pdf`,
and `.tex`. A filename without a directory writes to the current working
directory; a relative or full path writes elsewhere.

Plots use the same helper:

```{r plot-save, eval=FALSE}
p <- plot_compare(mtcars, variable = mpg, group = am, show_p = TRUE)
save_output(p, "mpg-by-transmission.png", width = 7, height = 5, dpi = 300)
```

Several tables and plots can be written to one Word report without calling
`officer` directly. Supply a named list; its names become section headings:

```{r combined-word-report, eval=FALSE}
overview <- describe_data(mtcars)
p <- plot_compare(mtcars, variable = mpg, group = am)

save_output(
  list(
    "Dataset overview" = overview,
    "Table 1" = res,
    "Fuel economy by transmission" = p
  ),
  "complete-report.docx",
  title = "Vehicle analysis",
  page_break = TRUE
)
```

## Practical rule

- Build statistics first.
- Customise once, near the end of the pipeline.
- Keep flextable for Office output.
- Use `to_gt()` or `engine = "gt"` only when the destination calls for gt.
