---
title: "Styling and Exporting Tables"
description: >
  Apply themes, relabel columns and rows, and export to Word.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Styling and Exporting Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r load}
library(gtstats)
```

## Overview

Once you have built a table with the `summary_table()` workflow and rendered it with `tbl_stats()`, two functions handle the final presentation step:

- **`customise_table()`** — applies a theme, adds titles, and relabels columns, rows, and factor levels.
- **`to_flextable()`** — converts the table to a `flextable` object for Word output.

---

## customise_table()

`customise_table()` accepts either the original `gtstats` result or a
rendered `gt` table. For routine use, pass the result directly and let the
function render and style it in one step.

```{r base-table}
res <- summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p()

tbl <- tbl_stats(res)
```

### Themes

```{r journal}
customise_table(tbl, theme = "journal")
```

```{r minimal}
customise_table(tbl, theme = "minimal")
```

```{r compact}
customise_table(tbl, theme = "compact")
```

| Theme | Best for |
|---|---|
| `"default"` | Exploratory work, no strong preference |
| `"journal"` | Academic manuscripts, journal submission |
| `"classic"` | Reports with traditional table borders |
| `"minimal"` | Slide decks, dashboards, modern reports |
| `"compact"` | Dense tables where space is limited |

The font and table width can also be controlled directly:

```{r font-width}
customise_table(
  res,
  theme = "journal",
  font = "Arial",
  font_size = 11,
  width = 90
)
```

---

### Titles and notes

```{r titles}
customise_table(
  tbl,
  theme       = "journal",
  title       = "Table 1. Baseline characteristics",
  subtitle    = "Stratified by transmission type",
  source_note = "Data: mtcars (Motor Trend, 1974)."
)
```

---

### Relabelling columns

Supply a named character vector — `"internal_name" = "Display label"`. Column names in `gtstats` tables follow the pattern `"am = 1"`, `"am = 0"`, `"Overall"`, `"Level"`.

```{r col-labels}
customise_table(
  tbl,
  theme      = "journal",
  col_labels = c(
    "Level"   = "",
    "am = 1"  = "Manual (n = 13)",
    "am = 0"  = "Automatic (n = 19)",
    "Overall" = "All vehicles (n = 32)"
  )
)
```

---

### Relabelling rows

Supply a named vector where names are the current row label values and values are the replacement labels.

```{r row-labels}
customise_table(
  tbl,
  theme      = "journal",
  row_labels = c(
    "mpg"    = "Miles per gallon",
    "wt"     = "Weight (1 000 lbs)",
    "cyl"    = "Cylinders",
    "vs (1)" = "V-shaped engine, n (%)",
    "Total (N)" = "Total"
  )
)
```

---

### Relabelling factor levels

Use `level_labels` to rename values that appear in the `Level` column of the rendered table.

```{r level-labels}
customise_table(
  tbl,
  level_labels = c(
    "4" = "4-cylinder",
    "6" = "6-cylinder",
    "8" = "8-cylinder"
  )
)
```

---

### Accent colour

Apply a brand or journal colour to the table borders:

```{r accent}
customise_table(tbl, theme = "journal", accent_color = "#123B7A")
```

---

### Row striping

```{r stripe}
customise_table(tbl, row_striping = TRUE, stripe_color = "#F0F4F8")
```

---

### A fully styled table

```{r full-style}
summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p() |>
  tbl_stats() |>
  customise_table(
    theme        = "journal",
    title        = "Table 1. Baseline characteristics by transmission type",
    source_note  = "SD = standard deviation; IQR = interquartile range; CI = confidence interval.",
    col_labels   = c(
      "Level"   = "",
      "am = 1"  = "Manual",
      "am = 0"  = "Automatic",
      "Overall" = "All"
    ),
    row_labels   = c(
      "mpg"       = "Miles per gallon",
      "wt"        = "Weight (1 000 lbs)",
      "cyl"       = "Cylinders",
      "vs (1)"    = "V-shaped engine",
      "Total (N)" = "Total"
    ),
    level_labels = c(
      "4" = "4-cylinder",
      "6" = "6-cylinder",
      "8" = "8-cylinder"
    ),
    accent_color = "#123B7A"
  )
```

---

## to_flextable()

`to_flextable()` converts a `gt_desc_table` object — the object **before** `tbl_stats()` — into a `flextable`. This is the recommended path for Word output via R Markdown or Quarto.

```{r flextable}
res <- summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p()

ft <- to_flextable(
  res,
  font = "Arial",
  font_size = 10,
  show_footnotes = TRUE
)
ft
```

In an R Markdown document with `output: word_document`, simply print `ft` in a code chunk and it will render as a formatted table in the Word file.

## Saving tables and plots

gtstats provides simple helpers to export tables and plots for reports, presentations, and manuscripts.

### Save a table

You can save either a `gtstats` object or a rendered `gt_tbl`.

If a `gtstats` object is supplied, `save_output()` internally renders it
using `tbl_stats()`.

```{r}
res <- summary_table(mtcars, by = am, overall = TRUE) |>
  add_summary(vars = c(mpg, wt, cyl)) |>
  add_proportion(var = vs) |>
  add_total() |>
  add_p()
```

Save directly from the gtstats object:

```{r}
save_output(res, "table1.html")
```


Save from a pre-rendered gt object:

```{r}
gt_obj <- tbl_stats(res)

save_output(gt_obj, "table1.png")
```

Supported table formats are HTML, PNG, PDF, RTF, LaTeX (`.tex`), and DOCX.
A simple filename saves in the current working directory. A relative or full
filename can be supplied when the destination is elsewhere:

```{r}
save_output(res, "results/table1.docx")
```

### Save a plot

Plots created with `plot_compare()` can be saved using `save_output()`.

```{r}
p <- plot_compare(
  mtcars,
  outcome = mpg,
  by = am,
  show_p = TRUE
)

p
```


Save the plot:
```{r}
save_output(p, "mpg_by_am.png")
```

You can control output size and resolution:
```{r}
save_output(
  p,
  "mpg_by_am_highres.png",
  width = 7,
  height = 5,
  dpi = 300
)
```


---

## Tips

- `customise_table()` accepts either the original result or the `gt`
  object returned by `tbl_stats()`.
- `to_flextable()` works on the original `gtstats` object, **not** on the gt output. Call it before `tbl_stats()`.
- For the most control over Word output, combine `to_flextable()` with `flextable`'s own formatting functions.
